---
output:
  pdf_document: default
  html_document: default
---
```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(tidytext)
library(textstem)
library(clinspacy)
library(topicmodels)
library('reshape2')
library(stringr)

```

This practical is based on exploratory data analysis, named entity recognition, and topic modelling of unstructured medical note free-text data derived from electronic medical records (EMR).  Real EMR data is very difficult to access without a specific need/request so this data set is derived from [medical transcription](https://mtsamples.com/) data instead.  I'll also caveat that the options of natural language processing (NLP) in R are far inferior to those available in Python. 

First, install the packages in the setup block (`install.packages(c("readr", "dplyr", "tidyr", "ggplot2", "tidtext", "textstem", "clinspacy", "topicmodels", "reshape2"))`).

Note: To try and make it clearer which library certain functions are coming from clearer, I'll try to do explicit imports throughout this notebook.

## Data Parsing

After that we can grab the dataset directly from the `clinspacy` library.

```{r}
raw.data <- clinspacy::dataset_mtsamples()
dplyr::glimpse(raw.data)
```

There is no explanation or data dictionary with this dataset, which is a surprisingly common and frustrating turn of events!  

**1** Using the output of dplyr's `glimpse` command (or rstudio's data viewer by clicking on `raw.data` in the Environment pane) provide a description of what you think each in this dataset contains.

note_id is made to keep track of the amount of notes taken 

description is to describe the patient symptoms 

medical_specialty is what the patient has gone through for treatment that be surgery, urology and more 

sample_name is what they can give the patient to help with what they might have, or what they do to a patient to treat them that be hip revision or knee replace etc. 

transcription is a document that relates to the patient and doctor, to help the doctor make the right decision on medical treatment. 

keywords - words that were mention in the description, medical specialty, sample_name, and transcription. Words that were repeated in each section.

Let's see how many different medical specialties are featured in these notes: 
```{r}

raw.data %>% dplyr::select(medical_specialty) %>% dplyr::n_distinct()
```

So, how many transcripts are there from each specialty:


```{r}
ggplot2::ggplot(raw.data, ggplot2::aes(y=medical_specialty)) + ggplot2::geom_bar() + labs(x="Document Count", y="Medical Speciality")
```

Let's make our life easier and filter down to 3 specialties: a diagonstic/lab, a medical, and a surgical specialty

```{r} 
filtered.data <- raw.data %>% dplyr::filter(medical_specialty %in% c("Orthopedic", "Radiology", "Surgery")) 
```

## Text Processing

Let's now apply our standard pre-processing to the transcripts from these specialties.  
We are going to use the `tidytext` package to tokenise the transcript free-text.  
Let's remove stop words first. e.g., "the", "of", "to", and so forth.  These are known as stop words and we can remove them relative easily using a list from  `tidytext::stop_words` and `dplyr::anti_join()`

```{r}

analysis.data <- filtered.data %>%
  unnest_tokens(word, transcription) %>%
  mutate(word = str_replace_all(word, "[^[:alnum:]]", "")) %>%
  filter(!str_detect(word, "[0-9]")) %>%
  anti_join(stop_words) %>%
  group_by(note_id) %>%
  summarise(transcription = paste(word, collapse = " ")) %>%
  left_join(select(filtered.data, -transcription), by = "note_id")
```


Now let's tokenize the `transcription` to words (unigram) 
By default this tokenises to words but other options include characters, n-grams, sentences, lines, paragraphs, or separation around a regular expression.

```{r}
tokenized.data.unigram <- analysis.data %>% tidytext::unnest_tokens(word, transcription, to_lower=TRUE)
```

You can also do bi-grams
```{r}
tokenized.data <- analysis.data %>% tidytext::unnest_tokens(ngram, transcription, token = "ngrams", n=2, to_lower = TRUE)
```


**2** How many unique unigrams are there in the transcripts from each specialty:

3 unique, Surgrey 11977, Radiology 5935	and Orthopedic 7682	

```{r}
tokenized.data.unigram %>% dplyr::group_by(medical_specialty) %>% dplyr::distinct(word) %>% dplyr::summarise(n=dplyr::n())

```
For bigram

```{r}
tokenized.data %>% dplyr::group_by(medical_specialty) %>% dplyr::distinct(ngram) %>% dplyr::summarise(n=dplyr::n())
```

Let's plot some distribution of unigram tokens (words)


```{r}
word_counts <- tokenized.data.unigram %>%
    group_by(word) %>%
    summarise(count = n()) %>%
    ungroup() %>%
    arrange(desc(count))

count_distribution <- word_counts %>%
  group_by(count) %>%
  summarise(num_words = n()) %>%
  ungroup()
 
 ggplot2::ggplot(count_distribution, aes(x = count, y = num_words)) +
  geom_point() +
  labs(title = "Scatter Plot of Count Distribution",
       x = "Count of Unique Words",
       y = "Number of Words")
```


Let's plot some distribution of bigram tokens (words)


```{r}
word_counts <- tokenized.data %>%
    group_by(ngram) %>%
    summarise(count = n()) %>%
    ungroup() %>% 
    arrange(desc(count))

count_distribution <- word_counts %>%
  group_by(count) %>% 
  summarise(num_words = n()) %>% 
  ungroup()
 
 ggplot2::ggplot(count_distribution, aes(x = count, y = num_words)) +
  geom_point() +
  labs(title = "Scatter Plot of Count Distribution",
       x = "Count of Unique Bigrams",
       y = "Number of Words")
```

**3** How many unique bi-grams are there in each category without stop words and numbers?

Orthopedic	98129, Radiology	42857, Surgery	260849	

Used the above code under questions 2, to get the results removed the word distinct ungram
```{r}
tokenized.data.unigram %>% dplyr::group_by(medical_specialty) %>%  dplyr::summarise(n=dplyr::n())

```

Sometimes we are interested in tokenising/segmenting things other than words like whole sentences or paragraphs.  
**4** How many unique sentences are there in each category? Hint: use `?tidytext::unnest_tokens` to see the documentation for this function.

```{r}
uniq_sentence <- analysis.data %>% tidytext::unnest_tokens(ngram, transcription, token = "sentences", to_lower
 = TRUE)


uniq_sentence %>% dplyr::group_by(medical_specialty) %>% summarise(num_unique_sentences = n_distinct(uniq_sentence))
  
```
The number of unique sentences are  Orthopedic	1716, Radiology	1716, Surgery	1716			


Now that we've tokenized to words and removed stop words, we can find the most commonly word used within each category:

```{r}
tokenized.data %>%
  dplyr::group_by(medical_specialty) %>%
  dplyr::count(ngram, sort = TRUE) %>%
  dplyr::top_n(5)
```

We should lemmatize the tokenized words to prevent over counting of similar words before further analyses.  
Annoyingly, `tidytext` doesn't have a built-in lemmatizer.

**5** Do you think a general purpose lemmatizer will work well for medical data? Why not?


I don't think general purpose lemmatizer will work well, since I learned that alot of medical terminology will be interpret differently depending on the country and the language. From this we know that general purpose lemmatizer doesn't cover ever language and it might not be able to handle those medical terms and data from different country's.



Unfortunately, a specialised lemmatizer like in `clinspacy` is going to be very painful to install so we will just use a simple lemmatizer for now:

```{r}
lemmatized.data <- tokenized.data %>% dplyr::mutate(lemma=textstem::lemmatize_words(ngram))
```

We can now calculate the frequency of lemmas within each specialty and note.
```{r}
lemma.freq <- lemmatized.data %>% 
  dplyr::count(medical_specialty, lemma) %>%
  dplyr::group_by(medical_specialty) %>% 
  dplyr::mutate(proportion = n / sum(n)) %>%
  tidyr::pivot_wider(names_from = medical_specialty, values_from = proportion) %>%
  tidyr::pivot_longer(`Radiology`:`Surgery`,
               names_to = "medical_specialty", values_to = "proportion")
```

And plot the relative proportions 
```{r}

ggplot2::ggplot(lemma.freq, ggplot2::aes(x=proportion, 
                                         y=`Orthopedic`,
                                         color=abs(`Orthopedic` - proportion))) + 
  ggplot2::geom_abline(color="gray40", lty=2) +
  ggplot2::geom_jitter(alpha=0.1, size=2.5, width=0.3, height=0.3) +
  ggplot2::geom_text(ggplot2::aes(label=lemma), check_overlap=TRUE, vjust=1.5) +
  ggplot2::scale_x_log10(labels=scales::percent_format()) + 
  ggplot2::scale_y_log10(labels=scales::percent_format()) + 
  ggplot2::scale_color_gradient(limits=c(0, 0.001), low="darkslategray4", high="gray75") +
  ggplot2::facet_wrap(~medical_specialty, ncol = 2) +
  ggplot2::theme(legend.position="none") +
  ggplot2:: labs(y="Orthopedic", x = NULL)
```
**6** What does this plot tell you about the relative similarity of lemma frequencies between Surgery and Orthopedic and between radiology and Surgery? Based on what these specialties involve, is this what you would expect?

the relative similarity of lemma frequencies between Surgery and Orthopedic is that due to Orthopedic having so many terms compared to surgrey which has a diverse range of medical terms that related to surgerical procedures. 

For Radiology and Surgery we can see that the lemma frequencies for radiology would have more due to the terms related to imaging like "series images", while surgery has a wider range of terms. 
 
Based on the specialties involved this was actually what I did expected due to how different the term are for each specialty and how some have more terms than others.


**7** Modify the above plotting code to do a direct comparison of Surgery and Radiology (i.e., have Surgery or Radiology on the Y-axis and the other 2 specialties as the X facets)

COPIED THE ABOVE CODE AND MODIFIED 
```{r}
lemma.freq <- lemmatized.data %>% 
  dplyr::count(medical_specialty, lemma) %>%
  dplyr::group_by(medical_specialty) %>% 
  dplyr::mutate(proportion = n / sum(n)) %>%
  tidyr::pivot_wider(names_from = medical_specialty, values_from = proportion) %>%
  tidyr::pivot_longer(`Radiology`:`Orthopedic`,
               names_to = "medical_specialty", values_to = "proportion")
```



```{r}
ggplot2::ggplot(lemma.freq, ggplot2::aes(x=proportion, 
                                         y=`Surgery`,
                                         color=abs(`Surgery` - proportion))) + 
  ggplot2::geom_abline(color="gray40", lty=2) +
  ggplot2::geom_jitter(alpha=0.1, size=2.5, width=0.3, height=0.3) +
  ggplot2::geom_text(ggplot2::aes(label=lemma), check_overlap=TRUE, vjust=1.5) +
  ggplot2::scale_x_log10(labels=scales::percent_format()) + 
  ggplot2::scale_y_log10(labels=scales::percent_format()) + 
  ggplot2::scale_color_gradient(limits=c(0, 0.001), low="darkslategray4", high="gray75") +
  ggplot2::facet_wrap(~medical_specialty, ncol = 2) +
  ggplot2::theme(legend.position="none") +
  ggplot2:: labs(y="Surgery", x = NULL)
```

### TF-IDF Normalisation

Maybe looking at lemmas across all notes in a specialty is misleading, what if we look at lemma frequencies across a specialty.

```{r}
lemma.counts <- lemmatized.data %>% dplyr::count(medical_specialty, lemma)
total.counts <- lemma.counts %>% 
                      dplyr::group_by(medical_specialty) %>% 
                      dplyr::summarise(total=sum(n))

all.counts <- dplyr::left_join(lemma.counts, total.counts)
```
Now we can calculate the term frequency / invariant document frequency (tf-idf):

```{r}
all.counts.tfidf <- tidytext::bind_tf_idf(all.counts, lemma, medical_specialty, n) 
```

We can then look at the top 10 lemma by tf-idf within each specialty:

```{r}
all.counts.tfidf %>% dplyr::group_by(medical_specialty) %>% dplyr::slice_max(order_by=tf_idf, n=10)
```
**8** Are there any lemmas that stand out in these lists? Why?

Yes for example "range motion" lemma stand out due to its high tf score, and overall score across the bored compared to the rest of the lemma, reason for that could be that this lemma is more specific than the other and its used way more.

We can look at transcriptions using these unusual lemmas to check how they are used with `stringr::str_detect`
```{r}
analysis.data %>% dplyr::select(medical_specialty, transcription) %>% dplyr::filter(stringr::str_detect(transcription, 'b.i.d')) %>% dplyr::slice(1)
```

**9** Extract an example of one of the other unusual "top lemmas" by modifying the above code

COPIED THE ABOVE CODE TO MODIFY IT 
```{r}
analysis.data %>% dplyr::select(medical_specialty, transcription) %>% dplyr::filter(stringr::str_detect(transcription, 'block foot')) %>% dplyr::slice(1)
```
## Topic Modelling

In NLP, we often have collections of documents (in our case EMR transcriptions) that we’d like to divide into groups so that we can understand them separately. Topic modeling is a method for unsupervised classification of such documents, similar to clustering on numeric data.

Latent Dirichlet allocation (LDA) is a particularly popular method for fitting a topic model. It treats each document as a mixture of topics, and each topic as a mixture of words. This allows documents to “overlap” each other in terms of content, rather than being separated into discrete groups, in a way that mirrors typical use of natural language.


- Every document is a mixture of topics. We imagine that each document may contain words from several topics in particular proportions. For example, in a two-topic model we could say “Document 1 is 90% topic A and 10% topic B, while Document 2 is 30% topic A and 70% topic B.”


- Every topic is a mixture of words. For example, we could imagine a two-topic model of American news, with one topic for “politics” and one for “entertainment.” The most common words in the politics topic might be “President”, “Congress”, and “government”, while the entertainment topic may be made up of words such as “movies”, “television”, and “actor”. Importantly, words can be shared between topics; a word like “budget” might appear in both equally.

LDA is a mathematical method for estimating both of these at the same time: finding the mixture of words that is associated with each topic, while also determining the mixture of topics that describes each document. There are a number of existing implementations of this algorithm, and we’ll explore one of them in depth.

First lets calculate a term frequency matrix for each transcription:
```{r}

lemma.counts <- lemmatized.data %>% dplyr::count(note_id, lemma)
total.counts <- lemma.counts %>% 
                      dplyr::group_by(note_id) %>% 
                      dplyr::summarise(total=sum(n))

all.counts <- dplyr::left_join(lemma.counts, total.counts)

emr.dcm <- all.counts %>% tidytext::cast_dtm(note_id, lemma, n)
```

Then we can use LDA function to fit a 5 topic (`k=5`) LDA-model
```{r}
emr.lda <- topicmodels::LDA(emr.dcm, k=6, control=list(seed=42))
emr.topics <- tidytext::tidy(emr.lda, matrix='beta')
```

Then we can extract the top terms per assigned topic:
```{r}

top.terms <- emr.topics %>% dplyr::group_by(topic) %>% 
  dplyr::slice_max(beta, n=10) %>%
  dplyr::ungroup() %>%
  dplyr::arrange(topic, -beta)


top.terms %>% 
  dplyr::mutate(term=tidytext::reorder_within(term, beta, topic)) %>% 
  ggplot2::ggplot(ggplot2::aes(beta, term, fill=factor(topic))) + 
    ggplot2::geom_col(show.legend=FALSE) + 
    ggplot2::facet_wrap(~ topic, scales='free')  +
    tidytext::scale_y_reordered()
```



Now we can ask how well do these assigned topics match up to the medical specialties from which each of these transcripts was derived.

```{r}
specialty_gamma <- tidytext::tidy(emr.lda, matrix='gamma')

# we need to join in the specialty from the note_id
note_id_specialty_mapping <- lemmatized.data %>%
  dplyr::mutate(document=as.character(note_id)) %>% 
  dplyr::select(document, medical_specialty) %>% 
  dplyr::distinct()

specialty_gamma <- dplyr::left_join(specialty_gamma, note_id_specialty_mapping)
```

```{r}

specialty_gamma %>%
  dplyr::mutate(medical_specialty = reorder(medical_specialty, gamma * topic)) %>%
  ggplot2::ggplot(ggplot2::aes(factor(topic), gamma)) +
  ggplot2::geom_boxplot() +
  ggplot2::facet_wrap(~ medical_specialty) +
  ggplot2::labs(x = "topic", y = expression(gamma))
```

Interestingly, Surgery assigns mostly to a single topic but radiology and Orthopedic are both more diverse in transcriptions.  We'd possibly expect this from radiology due to referring to imaging for many different diagnoses/reasons. 
However, this may all just reflect we are using too few topics in our LDA to capture the range of possible assignments. 

**10** Repeat this with a 6 topic LDA, do the top terms from the 3 topic LDA still turn up? How do the specialties get split into sub-topics?

I changed the above code on line 320 with 6 topic LDA. Some of the top terms do show up on the 3 topic LDA, plus
the specialties get split into sub-topic due to the relationship  found in the medical data regarding related words that are than assigns to a sub-topic. These related words that are usually seen together are found and reused in different specialties.   


## Credits

Examples draw heavily on material (and directly quotes/copies text) from Julia Slige's `tidytext` textbook.


